\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-10 k-planar.tex
cleaned up  proof in appendix, still need to think more about when k \neq 3

For a cycle $C$, we write $C = [ P_1 , \dots , P_k ]$ if $P_1 , \dots, P_k$ are pairwise disjoint non-empty paths in $C$, and the endpoints of each path $P_i$ can be labelled $x_i$ and $y_i$ so that $y_ix_{i+1} \in E(C)$ for $i\in\{1,\dots,k\}$, where $x_{k+1}$ means
$x_1$. This implies that $\{P_1,\dots,P_k\}$ is a vertex-partition of $C$.

@@

@@

Say $G$ is a 2-connected plane multigraph. Let $T$ be a spanning tree of $G$ rooted at a vertex $r$ on the outerface of $G$. A \defin{tripod} (with respect to $T$) is any subgraph of $G$ consisting of three vertical paths $P,Q,R$ in $T$ plus a facial cycle $F$ of $G$,
where the lower endpoint of each of $P,Q,R$ is in $F$. A \defin{bipod} (with respect to $T$) is any (connected) subpath of a tripod (with respect to $T$).
 %is a path $PXQ$ where $P$ and $Q$ are vertical paths in $T$ and $X$ is a facial path of $G$, where the lower endpoints of $P$ and $Q$ are the endpoints of $X$. 
\note{DW}{Need to think about degenerate cases. We want every bipod to be a tripod}
%, and every connected subgraph of a bipod to be a bipod.}.
These definitions generalise definitions in \citep{DJMMUW20} where $G$ is assumed to be a plane triangulation, in which case $F$ is a 3-cycle and $X$ is an edge. \note{DW}{We actually managed to omit the definition of `bipod' in \citep{DJMMUW20}!}

@@

\begin{lem}
Say $G_0$ is a 2-connected plane multigraph. Let $T$ be a spanning tree of $G_0$ rooted at some vertex $r$ on the boundary of the outerface of $G_0$. Let $k\in\{1,2,3\}$ and let $C = [A_1,\dots,A_k]$ be a cycle in $G_0$ with $r$ in the exterior of $C$, where $A_1, \dots, A_k$ are bipods in $G_0$. Let $G$ be the plane graph consisting of all the edges and vertices of $G_0$ in $C$ and the interior of $C$. Then $G$ has a partition $\PP$ into tripods with respect to $T$ such that:
\begin{compactitem}
\item $P_1,\dots,P_k\in\PP$,
\item $G / \PP$ has a tree-decomposition $\mathcal{T}$ of width at most 3 such that:
\begin{compactitem}
	\item some bag of $\mathcal{T}$ contains $P_1,\dots,P_k$, and 
	\item for every face $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ such that $V(F)\subseteq \bigcup_{X\in B} X$.
\end{compactitem}
\end{compactitem}
\end{lem}

\begin{proof}
	\note{DW}{Can we always assume $k=3$ by allowing $A_i=\emptyset$? That would clean up the proof.}

	\note{DW}{Is $C$ allowed to be a 2-cycle?}
The proof is by induction on $|V(G)\setminus V(C)|$. If $V(G)\setminus V(C)=\emptyset$, then $G$ has just one face bounded by $C$ and the desired partition is $\PP=\{P_1,\ldots,P_k\}$.  The tree-decomposition of $G/\PP$ consists of a single bag that contains the $k\le 3$ vertices corresponding to $P_1,\ldots,P_k$.

Now assume that $V(G)\setminus V(C)\neq\emptyset$. Partition the vertices of $C$ into paths $R_1$, $R_2$, and $R_3$ as follows:
\begin{compactenum}
	\item If $k=1$ then, since $C$ is a cycle, $P_1$ has at least three vertices, so $P_1=[v, P_1', w]$ for two distinct vertices $v$ and $w$.   \note{DW}{need to be careful with parallel edges} We set $R_1:=v$, $R_2:=P_1'$ and $R_3:=w$.

	\item If $k=2$ then we may assume without loss of generality that $P_1$ has at least two vertices so $P_1=[v,P_1']$ \note{DW}{need to be careful with parallel edges}.  We set $R_1:=v$, $R_2:=P_1'$ and $R_3:=P_2$.

@@

Let $G'$ be a near-triangulation obtained from $G$ by triangulating every internal face. For $i\in\{1,2,3\}$, colour each vertex in $R_i$ by $i$. Now, for each remaining vertex $v$ in $G$, consider the path $P_v$ from $v$ to the root of $T$. Since $r$ is on the outer-face of $G_0$, $P_v$ contains at least one vertex of $C$. If the first vertex of $P_v$ that belongs to $F$ is in $R_i$ then assign the colour $i$ to $v$. We obtain a 3-colouring of the vertices of $G'$ that satisfies the conditions of Sperner's Lemma. By Sperner's Lemma there exists a triangular face $\tau=v_1v_2v_3$ of $G'$ whose vertices are coloured $1,2,3$ respectively. Let $F$ be the face of $G_0$ containing $\tau$.

@@

Let $Y$ be the tripod consisting of $Q_1'\cup Q_2' \cup Q_3'$ plus the vertices and edges of $F$.

Say $F=v_1 S_1 v_2 S_2 v_3 S_3$, where $S_i$ is a facial path in $F$ clockwise from $v_i$ to $v_{i+1}$.

Let $L_i$ be the path $Q_i'S_iQ_{i+1}'$ for $i\in\{1,\dots,k\}$. So $L_i$ is a bipod.

, where we use the convention that $Q_4=Q_1$ and $R_4=R_1$.

@@

Consider the plane graph $Z$ consisting of the edges and vertices of $F$, the edges and vertices of $C$, and the edges and vertices of $Q_1,Q_2,Q_3$. Then $Z$ is 2-connected, each face of $Z$ is bounded by a cycle, and the outerface of $Z$ is $C$. Let $Z_1,\dots,Z_m$ be the internal faces of $Z$.

Consider $j\in\{1,\dots,m\}$. Let $G_j$ be the subgraph of $G$ consisting of all the edges and vertices of $G$ in $Z_j$ and the interior of $Z_j$. The outerface of $G_j$ is $Z_j$. Observe that for some $i\in\{1,\dots,k\}$, we have
$Z_j=[R_{i,j}^+,R_{i+1,j}^-,S_{i,j}]$, where 
$R_{i,j}^+$ is some subpath of $R_i^+$,
$R_{i+1,j}^-$ is some subpath of $R_{i+1}^-$, and
$S_{i,j}$ is some subpath of $S_i$.
Since $R_i^+$, $R_{i+1}^-$ and $S_i$ are bipods, 
$R_{i,j}^+$, $R_{i+1,j}^-$ and $S_{i,j}$ are bipods. 
By construction, $v_{i-1}$ is not in $G_j$. 
Thus $V(G_j)\setminus V(Z_j) \subsetneq V(G)\setminus V(C)$, and we may apply induction to $G_j$.  Thus $G_j$ has a partition $\PP_j$ into tripods with respect to $T$ such that:
\begin{compactitem}
	\item $R_{i,j}^+, R_{i+1,j}^-, S_{i,j}\in\PP_j$,
	\item the quotient $G_j / \PP_j$ has a tree-decomposition $\mathcal{T}_j=\{B_x:x\in V(D_j)\}$ of width at most 3 such that: 
	\begin{compactitem}
		\item some bag of $\mathcal{T}_j$ contains $R_{i,j}^+, R_{i+1,j}^-, S_{i,j}$, and 
		\item for every face $F$ of $G_j$ there is a bag $B$ of $\mathcal{T}_j$ such that $V(F)\subseteq \bigcup_{X\in B} X$.
	\end{compactitem}
\end{compactitem}

%This graph has up to three \note{DW}{can be more} internal faces $C_1,C_2,C_3$ where each $C_i=[Q_i',R_i^+,R_{i+1}^-,Q_{i+1}']$ and $R_{i}^+$ and $R_{i}^-$ are the same portions of $R_i$ as defined in \cref{NearTriang} \note{DW}{move in here}    Observe that  $C_i=[ R_i^+ , R_{i+1}^-, I_i ]$, where $R_i^+$ and $R_{i+1}^-$ are bipods, and $I_i$ is the bipod $Q_i' S_i Q_{i+1}'$. Let $G_i$ be the subgraph of $G$ whose vertices and edges are in $C_i$ or its interior.

@@

We now construct the desired partition $\PP$ of $G$. 
Let $\PP$ be obtained from $\bigcup_{j\in\{1,\dots,m\}} \PP_j$ by replacing 
$R_{i,j}^+\in\PP_j$ by $P_i$, 
$R_{i+1,j}^-\in\PP_j$ by $P_i$, and
$S_{i,j}\in\PP_j$ by $Y$.
By construction, $\PP$ partitions $V(G)$ into tripods with respect to $T$ and $P_1,\ldots,P_k\in \PP$.

We now construct the desired tree-decomposition of $G/\PP$. For each $j\in\{1,\dots,m\}$, apply the above replacements within the bags of $\mathcal{T}_j$, and let $u_j$ be the node of $D_j$ such that $R_{i,j}^+, R_{i+1,j}^-, S_{i,j} \in B_{u_j}$. Let $D$ be the tree obtained from the disjoint union of $D_1,\dots,D_m$ by adding one new node $u$ adjacent to $u_1,\dots,u_m$. Define $B_u:=\{P_1,P_2,P_3,Y\}$. We claim that $\{B_x:x\in V(D)\}$ is the desired tree-decomposition of $G/\PP$. By construction, every bag $B_x$ has size at most 4.

First we show that for each part $X\in\PP$, the set $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$. If $X\not\in \{P_1,P_2,P_3,Y\}$ then $X$ is only in bags in some $D_j$, implying $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$. If $X\in \{P_1,P_2,P_3,Y\}$ then $X$ is in $B_u$, and if some part $X'$ of some $\PP_j$ was replaced by $X$ in the above construction, then $X'$ is in $B_{u_j}$ and $\{x\in V(D_j):X\in B_x\}$ induces a subtree of $D_j$, implying that  $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$ (since $uu_j\in E(D)$).

We now show that, for every edge $XW$ of $G/\PP$, there is a node $x\in V(D)$ such that $X,W\in B_x$. If $X,W\in \{P_1,P_2,P_3,Y\}$ then $X,W\in B_u$ as desired. Otherwise, without loss of generality, $X\in \PP_j\setminus \{P_1,P_2,P_3,Y\}$ for some $j\in\{1,\dots,m\}$. 
Say $X'\in\PP'_j$ replaced by $X\in\PP_j$. Thus $X\subseteq V(G_j)\setminus V(Z_j)$. Since $Z_j$ separates $V(G_j)\setminus V(Z_j)$ and $V(G)\setminus V(G_j)$ and $XW\in E(G/\PP)$, it must be that $W\in \PP_j$ also, implying that $XW\in B_x$ for some node $x\in V(D_j)$. 
This concludes the proof that $(B_x:x\in V(D))$ is the desired tree-decomposition  of $G/\PP$.

Every face $F$ of $G$ is a face of some $G_j$ or is $F$ \note{DW}{Avoid using $F$ for two different things}. In the latter case, if $B$ is the bag at $u$ then $V(F)\subseteq \bigcup_{X\in B} X$ since $V(F) \subseteq Y\in B$. In the first case, there is a bag $B$ of $\mathcal{T}_j$ such that $V(F)\subseteq \bigcup_{X\in B} X$. This property is maintained by the above replacements since 
$R_{i,j}^+ \subseteq P_i$, 
$R_{i+1,j}^- \subseteq P_i$, and
$S_{i,j}\subseteq Y$.

@@

This completes the description of $(B_x:x\in V(J))$. Clearly, every bag $B_x$ has size at most 4. We now prove that $(B_x:x\in V(J))$  is a tree-decomposition of $H$ 
\end{proof}

>>>>

For a cycle $C$, we write $C = [ P_1 , \dots , P_k ]$ if $P_1 , \dots, P_k$ are pairwise disjoint non-empty paths in $C$, and the endpoints of each path $P_i$ can be labelled $x_i$ and $y_i$ so that $y_ix_{i+1} \in E(C)$ for $i\in\{1,\dots,k\}$, where $x_{k+1}$ means
$x_1$. This implies that $\{P_1,\dots,P_k\}$ is a vertex-partition of $C$. \note{DW}{Allow $P_i$ to be empty?}

@@

\note{DW}{It would help to say that we consider a path to be an (unordered) subgraph. Would that be a problem anywhere else?}

@@

Say $G$ is a 2-connected plane multigraph. Let $T$ be a spanning tree of $G$ rooted at a vertex $r$ on the outerface of $G$. A \defin{bipod} in $G$ (with respect to $T$) is any path $P_1SP_2$, where $S$ is a facial path in $G$, and each $P_i$ is a vertical path in $T$ whose lower endpoint is in $S$. Here it is possible that $S$ or $P_i$ is empty. So every subpath of a bipod is a bipod. A \defin{tripod} in $G$ (with respect to $T$) is either a bipod in $G$ or a subgraph of $G$ consisting of a facial cycle $F$ of $G$, plus three distinct vertical paths $P_1,P_2,P_3$ in $T$ where the lower endpoint of each $P_i$ is in $F$. \note{DW}{Need to think about degenerate cases.}. These definitions generalise definitions in \citep{DJMMUW20} where $G$ is assumed to be a plane triangulation, in which case $F$ is a 3-cycle. \note{DW}{We actually managed to omit the definition of `bipod' in \citep{DJMMUW20}!}

@@

\begin{lem}
Let $G_0$ be a 2-connected plane multigraph. Let $T$ be a spanning tree of $G_0$ rooted at some vertex $r$ on the boundary of the outerface of $G_0$. Let $k\in\{1,2,3\}$ and let $C = [P_1,\dots,P_k]$ be a cycle in $G_0$ with $r$ in the exterior of $C$, where $P_1, \dots, P_k$ are bipods in $G_0$. 
	\note{DW}{We need to assume that $|V(C)|\geq 3$ and ensure that when we apply induction $|V(Z_j)|\geq 3$. I think it is okay, but it needs an argument.}
	Let $G$ be the plane graph consisting of all the edges and vertices of $G_0$ in $C$ and the interior of $C$. Then $G$ has a partition $\PP$ into tripods with respect to $T$ such that:
\begin{compactitem}
\item $P_1,\dots,P_k\in\PP$,
\item $G / \PP$ has a tree-decomposition $\mathcal{T}$ of width at most 3 such that:
\begin{compactitem}
	\item some bag of $\mathcal{T}$ contains $P_1,\dots,P_k$, and 
	\item for every facial cycle $F$ of $G$ there is a bag $B$ of $\mathcal{T}$ such that $V(F)\subseteq \bigcup_{X\in B} X$.
\end{compactitem}
\end{compactitem}
\end{lem}

\begin{proof}
\note{DW}{Can we assume $k=3$ by allowing $P_i=\emptyset$? That would clean up the proof. We could scrap $R_1,R_2,R_3$ and just use $P_1,P_2,P_3$.}

The proof is by induction on $|V(G)\setminus V(C)|$. If $V(G)\setminus V(C)=\emptyset$, then $G$ has just one face bounded by $C$ and the desired partition is $\PP=\{P_1,\ldots,P_k\}$.  The desired tree-decomposition of $G/\PP$ consists of a single bag $\{P_1,\ldots,P_k\}$ indexed by a 1-node tree.

Now assume that $V(G)\setminus V(C)\neq\emptyset$. Partition the vertices of $C$ into paths $R_1$, $R_2$, and $R_3$ as follows:
\begin{compactenum}
	\item If $k=1$ then, since $C$ is a cycle, $P_1$ has at least three vertices, so $P_1=[v, P_1', w]$ for two distinct vertices $v$ and $w$. We set $R_1:=v$, $R_2:=P_1'$ and $R_3:=w$.

	\item If $k=2$ then we may assume without loss of generality that $P_1$ has at least two vertices so $P_1=[v,P_1']$. We set $R_1:=v$, $R_2:=P_1'$ and $R_3:=P_2$.

@@

Let $G'$ be a near-triangulation obtained from $G$ by triangulating every internal face. For $i\in\{1,2,3\}$, colour each vertex in $R_i$ by $i$. Now, for each remaining vertex $v$ in $G$, consider the path $P_v$ from $v$ to the root of $T$. Since $r$ is on the outer-face of $G_0$, $P_v$ contains at least one vertex of $C$. If the first vertex of $P_v$ that belongs to $F$ is in $R_i$, then assign the colour $i$ to $v$. We obtain a 3-colouring of the vertices of $G'$ that satisfies the conditions of Sperner's Lemma. By Sperner's Lemma there exists a triangular face $\tau=v_1v_2v_3$ of $G'$ whose vertices are coloured $1,2,3$ respectively. Let $F$ be the cycle bounding the face of $G$ containing $\tau$.

@@

Let $Y$ be the tripod $F\cup Q_1'\cup Q_2' \cup Q_3'$. Say $F=v_1 S_1 v_2 S_2 v_3 S_3$, where $S_i$ is a facial path in $F$ clockwise from $v_i$ to $v_{i+1}$ (where $v_4=v_1$). Let $L_i$ be the path $Q_i'S_iQ_{i+1}'$ for $i\in\{1,2,3\}$ (where $Q_4=Q_1$ and $R_4=R_1$). So $L_i$ is a bipod. \note{DW}{Note that $S_i\neq\emptyset$ since $v_i\neq v_{i+1}$.}

@@

Let $Z$ be the plane graph $F\cup C\cup Q_1\cup Q_2\cup Q_3$. Then $Z$ is 2-connected, each face of $Z$ is bounded by a cycle, and the outerface of $Z$ is $C$. Let $Z_1,\dots,Z_m$ be the internal facial cycles of $Z$ except $F$.

\note{DW}{For a cycle $C$ in a plane graph $G$, define $G\langle C\rangle$ to be the subgraph of $G$ consisting of $C$ and all the edges and vertices of $G$ in the interior of $C$?}

Consider $j\in\{1,\dots,m\}$. Let $G_j$ be the subgraph of $G$ consisting of $Z_j$ and all the edges and vertices of $G$ in the interior of $Z_j$. The outerface of $G_j$ is $Z_j$. Observe that for some $i\in\{1,2,3\}$, we have
$Z_j=[R_{i,j}^+,R_{i+1,j}^-,S_{i,j}]$, where 
$R_{i,j}^+$ is some subpath of $R_i^+$,
$R_{i+1,j}^-$ is some subpath of $R_{i+1}^-$, and
$L_{i,j}$ is the path $Z_j-V(C)$ (which is contained in $Y$). 
Since $R_i^+$, $R_{i+1}^-$ and $L_i$ are bipods, 
$R_{i,j}^+$, $R_{i+1,j}^-$ and $L_{i,j}$ are bipods. 
By construction, $v_{i-1}$ is not in $G_j$ (where $v_0=v_3$). 
Thus $V(G_j)\setminus V(Z_j) \subsetneq V(G)\setminus V(C)$, and we may apply induction to $G_j$.  Thus $G_j$ has a partition $\PP_j$ into tripods with respect to $T$ such that:
\begin{compactitem}
	\item $R_{i,j}^+, R_{i+1,j}^-, L_{i,j}\in\PP_j$, 
	\item the quotient $G_j / \PP_j$ has a tree-decomposition $\mathcal{T}_j=\{B_x:x\in V(D_j)\}$ of width at most 3 such that: 
	\begin{compactitem}
		\item some bag $B_{u_j}$ of $\mathcal{T}_j$ contains $R_{i,j}^+, R_{i+1,j}^-, L_{i,j}$, and 
		\item for every facial cycle $F'$ of $G_j$ there is a bag $B$ of $\mathcal{T}_j$ such that $V(F')\subseteq \bigcup_{X\in B} X$.
	\end{compactitem}
\end{compactitem}

%This graph has up to three \note{DW}{can be more} internal faces $C_1,C_2,C_3$ where each $C_i=[Q_i',R_i^+,R_{i+1}^-,Q_{i+1}']$ and $R_{i}^+$ and $R_{i}^-$ are the same portions of $R_i$ as defined in \cref{NearTriang} \note{DW}{move in here}    Observe that  $C_i=[ R_i^+ , R_{i+1}^-, I_i ]$, where $R_i^+$ and $R_{i+1}^-$ are bipods, and $I_i$ is the bipod $Q_i' L_i Q_{i+1}'$. Let $G_i$ be the subgraph of $G$ whose vertices and edges are in $C_i$ or its interior.

@@

We now construct the desired partition $\PP$ of $G$. 
Let $\PP$ be obtained from $\bigcup_{j\in\{1,\dots,m\}} \PP_j$ by 
replacing $R_{i,j}^+\in\PP_j$ by $P_i$, 
replacing $R_{i+1,j}^-\in\PP_j$ by $P_{i+1}$, and
replacing $L_{i,j}\in\PP_j$ by $Y$. \note{DW}{What if $k\neq 3$? Then $P_2$ or $P_3$ might not be defined.}
By construction, $\PP$ partitions $V(G)$ into tripods with respect to $T$ and $P_1,\ldots,P_k\in \PP$.

We now construct the desired tree-decomposition of $G/\PP$. Let $D$ be the tree obtained from the disjoint union of $D_1,\dots,D_m$ by adding one new node $u$ adjacent to $u_1,\dots,u_m$. Define $B_u:=\{P_1,\dots,P_k,Y\}$. Apply the above replacements within the bags of $\mathcal{T}_1,\dots,\mathcal{T}_m$. We claim that $\mathcal{T}:=\{B_x:x\in V(D)\}$ is the desired tree-decomposition of $G/\PP$. By construction, every bag of $\mathcal{T}$ has size at most 4.

First we show that for each part $X\in\PP$, the set $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$. If $X\not\in \{P_1,P_2,P_3,Y\}$ then $X$ is only in bags in some $D_j$, implying $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$. If $X\in \{P_1,P_2,P_3,Y\}$ then $X$ is in $B_u$, and if some part $X'$ of some $\PP_j$ was replaced by $X$ in the above construction, then $X'$ is in $B_{u_j}$ and $\{x\in V(D_j):X'\in B_x\}$ induces a subtree of $D_j$ (since $\mathcal{T}_j$ is a tree-decomposition of $G_j$). Thus $\{x\in V(D) : X\in B_x\}$ induces a subtree of $D$ (since $uu_j\in E(D)$).

We now show that, for every edge $XW$ of $G/\PP$, there is a node $x\in V(D)$ such that $X,W\in B_x$. If $X,W\in \{P_1,P_2,P_3,Y\}$ then $X,W\in B_u$ as desired. Otherwise, without loss of generality, $X\in \PP_j\setminus \{P_1,P_2,P_3,Y\}$ for some $j\in\{1,\dots,m\}$. 
Thus $X\subseteq V(G_j)\setminus V(Z_j)$. Since $Z_j$ separates $V(G_j)\setminus V(Z_j)$ and $V(G)\setminus V(G_j)$ and $XW\in E(G/\PP)$, it must be that $W$ is also in $\PP_j$. Say $W$ replaced $W'$ in $\PP_j$ (where possibly $W=W'$). Since $\mathcal{T}_j$ is a tree-decomposition of $G_j$, $XW'\in B_x$ for some node $x\in V(D_j)$, implying $XW\in B_x$. This concludes the proof that $(B_x:x\in V(D))$ is the desired tree-decomposition of $G/\PP$.

Every facial cycle $F_0$ of $G$ is a face of some $G_j$ or is $F$. In the latter case, $V(F)\subseteq Y \subseteq \bigcup_{X\in B_u} X$. In the first case, there is a bag $B_x$ of $\mathcal{T}_j$ such that $V(F)\subseteq \bigcup_{X\in B_x} X$. This property is maintained by the above replacements since 
$R_{i,j}^+ \subseteq P_i$, 
$R_{i+1,j}^- \subseteq P_{i+1}$, and
$L_{i,j}\subseteq Y$. \note{DW}{What if $k\neq 3$?}

@@

This shows that $(B_x:x\in V(J))$ is the desired tree-decomposition of $G/\PP$. 
\end{proof}
